Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2020-03-11 paper/results.tex
incorporating Adrian's suggestions

To explore the relationship between rotation period, effective temperature
(\teff ) and velocity dispersion, we first removed high and low velocity
outliers from the \mct\ sample by performing 3$\sigma$ sigma-clipping on the
\vb\ velocities.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars, although this did not change the overall
trends in the data.
% We also limited the sample to temperatures in the range 6000 K $<$ \teff\ $<$
% 3500 K to avoid biases caused by the selection function at the faint, cool
% end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
Finally, we removed stars with rotation periods shorter than the main
population of periods, since this area of the period-\teff\ diagram is
sparsely populated.
We removed these rapid rotators by cutting out stars with gyrochronal ages
less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
line of constant age in period-\teff, or period-color space.} traces the
bottom edge of the main population of rotation periods.
After these cuts, around 13,000 stars out of the original $\sim$34,000 were
included in the sample.

The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by the standard deviation of their
(\vb) velocities, where \sigmavb\ was calculated for groups of stars over a
grid in $\log_{10}$(period) and temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
We discuss this assumption further in section
\ref{sec:mass-dependent-heating}.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages because they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr in both top
    and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from][]{berger2020}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}
Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
rotation period across all temperatures, implying that rotation period
increases with age, as expected.
This result is insensitive to the choice of bin position and size.
Black lines show gyrochrones from the \citet{angus2019} gyrochronology model,
which projects the rotation-color relation of Praesepe to longer rotation
periods over time.
These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
At the youngest ages, these gyrochrones describe the data well: the palest
yellow (youngest) stars with the lowest velocity dispersions all fall close to
the 0.5 Gyr gyrochrone.
% Lines of constant age (isochrones) appear to follow the shape of the
% Praesepe-based gyrochronology model (black dotted line) at young ages.
% However, at older ages it appears that the relation between rotation period
% and \teff\ flattens out, until eventually rotation period {\it decreases} with
% decreasing effective temperature at a given age.
However, although the 0.5 Gyr and 1 Gyr gyrochrones also trace constant
velocity dispersion/age among the field stars, by 1.5 Gyr the gyrochrones
start to {\it cross} different velocity dispersion regimes.
For example, the 1.5 Gyr gyrochrone lies on top of stars with velocity
dispersions of around 10-11 kms$^{-1}$ at 5000-5500K and stars with $\sim$15
\kms\ velocity dispersions at 4000-4500K.
The gyrochrones older than 1.5 Gyr also cross a range of velocity dispersions.
If these were true isochrones they would follow lines of constant velocity
dispersion.
At ages older than around 1 Gyr, it appears that gyrochrones should have a
more flattened, or even inverted, shape in rotation period-\teff\ space than
these Praesepe-based models.
% For late K-dwarfs older than around 1 Gyr, it appears that lower-mass stars
% rotate more rapidly than higher-mass stars.
% This is the opposite of the trend that is observed in young open clusters such
% as Praesepe and the Hyades.

@@

There is a sharp gap in the population of rotation periods, which lies just
above the 1 Gyr gyrochrone in the upper panel of figure \ref{fig:vplot}, whose
origin is unknown and is the subject of much speculation \citep{mcquillan2014,
davenport2018, reinhold2019}.
The Praesepe-based model appears to be valid below the gap but not above.
Although this may be coincidental (and more data would be needed to confirm a
connection) the gap may indeed separate a young regime where stellar cores are
decoupled from their envelopes from an old regime where these layers are more
tightly coupled.
If so, this could indicate that the phenomenon responsible for changing the
shape of gyrochrones in rotation-\teff\ space is related to the phenomenon that
produces the gap.

@@

\begin{figure}
  \caption{
      Top: rotation period vs. effective temperate for stars in the \mct\
    sample, separated into three groups. Blue circles
      show stars with rotation periods longer than the
    period gap, orange squares show stars with rotation periods shorter than
    the gap, but longer than the lower edge of the main rotation period
    distribution, and green triangles show stars with rotation periods shorter
    than this lower edge.
    Stars were separated into these three groups using \citet{angus2019}
    gyrochronology models, with the scheme shown in the legend.
    Only stars cooler than 5000 K are plotted in
    the bottom panel in order to isolate populations above and below the
    period gap, which only extends up to temperatures of $\sim$4600 K.
    Bottom: the velocities of these groups of stars (in the direction of
    Galactic latitude, $b$) are shown as a function of rotation period.
    The black line indicates the velocity standard deviation as a function of
    period.
}
  \centering
    \includegraphics[width=1\textwidth]{gap}
\label{fig:gap}
\end{figure}

@@

The bottom panel of figure \ref{fig:gap} shows the velocity dispersions of
stars in the \mct\ sample, with stars subdivided into three groups: those that
rotate more quickly than the major rotation period population (green
triangles), those with rotation periods shorter than the gap (orange squares),
and those with rotation periods longer than the gap (blue circles).
Stars were separated into these three groups using \citet{angus2019}
gyrochronology model, according to the scheme shown in the legend.
Only stars cooler than 5000 K are included in the bottom panel in order to
isolate populations above and below the period gap, which only extends up to a
temperature of $\sim$4600 K in our sample \citep[Although][found that the gap
extends to temperatures as hot as 6000 K]{davenport2017}.
In general, velocity dispersion increases with rotation period because both
quantities increase with age.
There is a smooth transition in velocity dispersion between stars with
rotation periods below and above the gap (orange squares to blue circles),
suggesting that these groups are part of the same Galactic population.
Previously, only the overall velocity dispersions of all stars above and below
the gap have been compared, leading to the assumption that these groups belong
to two distinct populations \citep{mcquillan2014}.

Synchronized binaries are pairs of stars whose rotation periods are equal to
their orbital period.
Since synchronization appears to happen at rotation periods of 7 days or
shorter \citep{simonian2019}, and most isolated stars have rotation periods
longer than 7 days, the rotation periods of synchronized binaries are likely
to be {\it shorter} than they would be if they were isolated stars.
In other words, their rotation periods do not reflect their ages.
Synchronized binaries are therefore a source of contamination for
gyrochronology and should be removed from samples before performing a
gyrochronal age analysis.
The gyrochronal age of a synchronized binary is likely to be much younger than
the true age of the system.
For this reason, stars that rotate rapidly but appear kinematically old are
likely to be synchronized binaries.
In general, the velocity dispersions of stars with rotation periods shorter
than the lower edge of the rotation period distribution (green triangles in
figure \ref{fig:gap}) are
not significantly smaller than the, presumed older, orange-colored stars.
In addition, some of the most rapidly rotating stars sample have relatively
large velocities.
The large velocity dispersions of the most rapidly rotating stars in the \mct\
sample indicates that at least some fraction of these stars are likely to be
synchronized binaries.
% This is an indication that many of these are synchronized binaries.
% The rotation periods of stars in synchronized binaries are tidally locked to
% the orbital period of the binary system and since synchronized binaries have
% short orbital periods this can produce old, seemingly isolated stars, with
% short rotation periods.
% Given the strong correlation between binary fraction and primary mass
% \citep[\eg][]{price-whelan2020}, and the fact that M dwarfs take longer to
% transition from the `I` (rapidly rotating) sequence to the `C' (slowly
% rotating) sequence, it is likely that
It general, there appears to be an increased probability of stars with
rotation periods less than $\sim$10 days being synchronized binaries.
This result is in agreement with a recent study which found that a large
fraction of photometric binaries were rapid rotators, and the probability of a
star being a synchronized binary system substantially increased below rotation
periods of around 7 days \citep{simonian2019}.

\subsection{Validating \vb\ dispersion as an age proxy}
\label{sec:mass-dependent-heating}

There are two main reasons why \vb\ velocity dispersion may not be a good age
proxy.
Firstly, mass-dependent heating may act on the sample, meaning that velocity
dispersion depends on both age and mass, so cannot be interpreted as a simple
age proxy.
Secondly, since stars in the \kepler\ field have a range of galactic
latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
stars, and introduce a velocity bias for high latitude stars (which are more
likely to be cooler and older).
In this section we demonstrate that neither of these problems seem to be a
significant issue for our data.

In order to establish whether \sigmavb\ can be used as an age proxy, we
searched for signs of mass-dependent heating within the \kepler\ field.
Mass-dependent dynamical heating may result from lower-mass stars experiencing
greater velocity changes when gravitationally perturbed than more massive
stars.
It has not been unambiguously observed in the galactic disk because of the
strong anti-correlation between stellar mass and stellar age.
Less massive stars do indeed have larger velocity dispersions, however they
are also older on average.
This mass-age degeneracy is highly reduced in M dwarfs because their
main-sequence lifetimes are longer than the age of the Universe, and no
evidence for mass-dependent heating has previously been found in M dwarfs
\citep[\eg][]{faherty2009, newton2016}.

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyr and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because it only
includes stars with {\it detectable} rotation periods, and since lower-mass
stars stay active for longer it is likely that it contains a strong mass-age
correlation.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
We also eliminated photometric binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
figure \ref{fig:age_gradient}).
We then applied the quality cuts described above in section
\ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
outliers before calculating the standard deviation of stars in each bin.
These extreme velocity outliers may be very old late K and M dwarfs, or they
result from using \vb\ instead of \vz, which introduces additional velocity
scatter.

Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
function of effective temperature \footnote{calculated by transforming
dereddened \gaia\ colors using equation \ref{eq:curtis}.} for the K and M
\kepler\ dwarf sample.
Velocity dispersion very slightly {\it decreases} with decreasing temperature,
the opposite of the trend expected for mass-dependent heating, however the
slope is only inconsistent with zero at 1.3 $\sigma$.
This trend may be due to a selection bias: cooler stars are fainter and
therefore typically closer, with smaller heights above the galactic plane and
smaller velocities.
The essential point however, is that we do not see evidence for mass-dependent
heating acting on stars in the \kepler\ field, indicating that velocity
dispersion {\it can} be used as an age proxy (with the caveat that there is
still a chance, albeit a small one, that the opposing effects of the selection
function and mass-dependent heating are working to cancel each other out).
This analysis was performed using \vb\ but we also examined the {\it vertical}
velocities of the 537 stars in this sample with RV measurements.
Again, no evidence was found for mass-dependent heating: the slope of the
velocity dispersion-temperature relation was consistent with zero.
\begin{figure}
  \caption{
      Top: Stellar velocity (\vb) as a function of \teff\ for
      \kepler\ K and M dwarfs.
Vertical lines indicate different \teff-groupings used to calculate velocity
    dispersion.
Pink stars were not included in velocity dispersion calculations as they were
    either removed as outliers during a sigma clipping process, or they lie at
    the sparcely populated, extremely cool end of the temperature range.
    Velocity dispersion and \teff\ are slightly positively correlated, likely
    due to a brightness-related selection bias, indicating that mass-dependent
    heating does not significantly affect low-mass stars in the \kepler\
    field.
}
  \centering
    \includegraphics[width=1\textwidth]{vb_vs_teff}
\label{fig:vb_vs_teff}
\end{figure}

Having found no strong evidence for mass-dependent heating, we next tested
the validity of \vb\ as a proxy for \vz\ in more detail.
At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
of $b$, this equivalence becomes an approximation that grows noisier with $b$.
To test the validity of the \vb$\sim$\vz\ approximation over a range of
latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
(GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
We downloaded stars from four pointings in the \kepler\ field with galactic
latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
\vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
Galaxy.
We subtracted this offset and examined the residuals of the \vz\ -- \vb\
relationship to investigate the variance as a function of Galactic latitude
(shown in figure \ref{fig:vb_vz}).
We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
$, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
20$^\circ$.

% Figure \ref{fig:vb_vz} shows the differences between \vz\ and \vb\ velocities
% over different Galactic latitudes.
% Kernel Density Estimates (KDEs) are shown as solid black lines and Gaussian
% fits are shown as dashed blue lines.
% The \vb\-\vz\ residuals are close to Gaussian, with slightly heavy tails, and
% the variance increases with increasing Galactic latitude.
\begin{figure}
  \caption{
This figure demonstrates the variance in the relationship between \vb\ and
    \vz\ for stars in the \kepler\ field, based on the GUMS simulation.
The panels show a kernel density estimator (KDE) (black solid line) for
    the \vz -- \vb\ residuals of stars in the GUMS simulation at four
    different Galactic latitudes.
Blue dashed lines show Gaussian fits to these KDEs.
The distributions are close to Gaussian, with slightly heavy tails.
The standard deviations of the Gaussian fits increase with Galactic latitude.
This figure illustrates how using \vb\ instead of \vz\ artificially
    increases velocity dispersion, especially at high latitudes.
}
  \centering
    \includegraphics[width=1\textwidth]{vb_vz}
\label{fig:vb_vz}
\end{figure}

Since we are concerned with velocity {\it dispersions}, rather than velocities
themselves, we also compared the \vb\ and \vz\ velocity dispersions as a
function of temperature for
stars downloaded from the GUMS simulation.
For stars at galactic latitudes of 15\degrees\ or less, \sigmavb\ was
consistent with $\sigma_{v{\bf z}}$, within uncertainties, however, at higher
latitudes the two quantities became significantly different.
For this reason we proceeded by only including stars with galactic latitudes
less than 15\degrees\ in our analysis.
Although we find that the transformation between \vz\ and \vb\ does not {\it
strongly} affect our results, we cannot rule out the possibility that it
introduces systematic biases into the velocity dispersions we present here.
In \gaia\ DR3, RVs will be available for most stars in this sample, providing
an opportunity to validate (or correct) the results presented here, and to
work in action-space, rather than velocity-space.

% Since we use \vb, not \vz\ in our analysis, it is possible that the sample
% selection function could influence our results.
% For example, higher-mass stars tend to be younger and reside closer to the
% galactic disk mid-plane (low galactic latitudes).
% Since \vb\ is only similar to \vz\ at low latitudes, this could mean that
% lower-mass stars typically have greater velocity dispersions due to .
% \teff\ and galactic latitude, $b$,
% could result in
% Stars at higher latitudes have additional velocity components in the ${\bf x}$
% and ${\bf y}$ directions, which could increase \vb\ but not \vz.
% Again however, since the relationship between \sigmavb\ and \teff\ is
% positively, not negatively correlated for cool stars in the \kepler\ field,
% this effect is probably too small to influence our results.

% Figure \ref{fig:vb_vs_teff} indicates that mass-dependent heating does not
% strongly affect the \mct\ sample of rotating \kepler\ stars.
% For this reason, we assume that age difference is the major cause of velocity
% dispersion differences between groups.
% In other words, (\vb) velocity dispersion is a reliable age proxy for the
% \mct\ sample.

>>>>

% To explore the relationship between rotation period, effective temperature
% (\teff ) and velocity dispersion,
% % , we first removed high and low velocity
% % outliers from the \mct\ sample by performing 3$\sigma$ sigma-clipping on the
% % \vb\ velocities.
% % Without sigma-clipping, we found that a small number of high velocity
% % outliers at the low-temperature end of our sample substantially raised the
% % velocity dispersion for cooler stars, although this did not change the overall
% % trends in the data.
% % We also limited the sample to temperatures in the range 6000 K $<$ \teff\ $<$
% % 3500 K to avoid biases caused by the selection function at the faint, cool
% % end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
% we removed stars with rotation periods shorter than the main population of
% periods, since this area of the period-\teff\ diagram is sparsely populated.
% We removed these rapid rotators by cutting out stars with gyrochronal ages
% less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
% Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
% line of constant age in period-\teff, or period-color space.} traces the
% bottom edge of the main population of rotation periods.
% After these cuts, around 13,000 stars out of the original $\sim$34,000 were
% included in the sample.

To explore the relationship between rotation period, effective temperature
(\teff ) and velocity dispersion, we calculated \sigmavb \footnote{\sigmavb
was calculated as 1.5$\times$ the median absolute deviation, to mitigate
sensitivity to outliers.} for groups of stars with similar rotation periods
and temperatures, and presumed similar age.
The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by \sigmavb, where \sigmavb\ was
calculated for groups of stars over a grid in $\log_{10}$(period) and
temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
We discuss this assumption further in the appendix.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages because they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr in both top
    and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from][]{berger2020}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}

Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
rotation period across all temperatures, implying that rotation period
increases with age, as expected.
This result is insensitive to the choice of bin position and size.
Black lines show gyrochrones from the \citet{angus2019} gyrochronology model,
which projects the rotation-color relation of Praesepe to longer rotation
periods over time.
These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
At the youngest ages, these gyrochrones describe the data well: the palest
yellow (youngest) stars with the lowest velocity dispersions all fall close to
the 0.5 Gyr gyrochrone.
% Lines of constant age (isochrones) appear to follow the shape of the
% Praesepe-based gyrochronology model (black dotted line) at young ages.
% However, at older ages it appears that the relation between rotation period
% and \teff\ flattens out, until eventually rotation period {\it decreases} with
% decreasing effective temperature at a given age.
However, although the 0.5 Gyr and 1 Gyr gyrochrones also trace constant
velocity dispersion/age among the field stars, by 1.5 Gyr the gyrochrones
start to {\it cross} different velocity dispersion regimes.
For example, the 1.5 Gyr gyrochrone lies on top of stars with velocity
dispersions of around 10-11 kms$^{-1}$ at 5000-5500K and stars with $\sim$15
\kms\ velocity dispersions at 4000-4500K.
The gyrochrones older than 1.5 Gyr also cross a range of velocity dispersions.
If these were true isochrones they would follow lines of constant velocity
dispersion.
At ages older than around 1 Gyr, it appears that gyrochrones should have a
more flattened, or even inverted, shape in rotation period-\teff\ space than
these Praesepe-based models.
% For late K-dwarfs older than around 1 Gyr, it appears that lower-mass stars
% rotate more rapidly than higher-mass stars.
% This is the opposite of the trend that is observed in young open clusters such
% as Praesepe and the Hyades.

@@

@@

\begin{figure}
  \caption{
      Top: rotation period vs. effective temperature for stars in the \mct\
    sample, separated into three groups. Blue circles
      show stars with rotation periods longer than the
    period gap, orange squares show stars with rotation periods shorter than
    the gap, but longer than the lower edge of the main rotation period
    distribution, and green triangles show stars with rotation periods shorter
    than this lower edge.
    Stars were separated into these three groups using \citet{angus2019}
    gyrochronology models, with the scheme shown in the legend.
    Only stars cooler than 5000 K are plotted in
    the bottom panel in order to isolate populations above and below the
    period gap, which only extends up to temperatures of $\sim$4600 K.
    Bottom: the velocities of these groups of stars (in the direction of
    Galactic latitude, $b$) are shown as a function of rotation period.
    The black line indicates the velocity standard deviation as a function of
    period.
}
  \centering
    \includegraphics[width=1\textwidth]{gap}
\label{fig:gap}
\end{figure}

@@

There is a sharp gap in the population of rotation periods (often called the
\kepler\ period gap), which lies just above the 1 Gyr gyrochrone in the upper
panel of figure \ref{fig:vplot}, whose origin is unknown and is the subject of
much speculation \citep{mcquillan2014, davenport2018, reinhold2019}.
This gap was first identified by \mct, and roughly follows a line of constant
gyrochronal age of around 1.1 Gyr \citep[according to the][gyrochronology
relation]{angus2019}.
Several explanations for the gap's origin have been proposed, including a
discontinuous star formation history \citep{mcquillan2013, davenport2017,
davenport2018} and a change in magnetic field structure causing a brief period
where rotational variability is reduced and rotation periods cannot be
measured \citep{reinhold2019}.

The top panel of figure \ref{fig:vplot} suggests that the \citet{angus2019},
Praesepe-based gyrochronology model is valid below the gap but not above.
Gyrochrones follow lines of constant velocity dispersion below the gap, but
{\it cross} lines of constant velocity dispersion above the gap.
Although we do not provide an in-depth analysis here (and more data may be
needed to confirm a connection) these data suggest that the gap may indeed
separate a young regime where stellar cores are decoupled from their envelopes
from an old regime where these layers are more tightly coupled.
If so, this could indicate that the phenomenon responsible for changing the
shape of gyrochrones in rotation-\teff\ space is related to the phenomenon that
produces the gap.

An alternate explanation for the gap is that the \mct\ sample contains two
distinct stellar populations: one young and one old.
If so, the kinematic properties of stars above and below the gap are likely to
be distinctly different.
The bottom panel of figure \ref{fig:gap} shows the velocity dispersions of
stars in the \mct\ sample, with stars subdivided into three groups: those that
rotate more quickly than the main rotation period population (green
triangles), those with rotation periods shorter than the gap (orange squares),
and those with rotation periods longer than the gap (blue circles).
Stars were separated into these three groups using \citet{angus2019}
gyrochronology model, according to the scheme shown in the legend.
Only stars cooler than 5000 K are included in the bottom panel in order to
isolate populations above and below the period gap, which only extends up to a
temperature of $\sim$4600 K in our sample \citep[Although][found that the gap
extends to temperatures as hot as 6000 K]{davenport2017}.
In general, velocity dispersion increases with rotation period because both
quantities increase with age.
There is a smooth transition in velocity dispersion between stars with
rotation periods below and above the gap (orange squares to blue circles),
suggesting that these groups are part of the same Galactic population.
Previously, only the overall velocity dispersions of all stars above and below
the gap have been compared, leading to the assumption that these groups belong
to two distinct populations \citep{mcquillan2014}.
The smooth increase in velocity dispersion across the gap shown here supports
the hypothesis that the gap is caused by a phenomenon of magnetic or
structural stellar evolution, not a discontinuity in the local star formation
history.

Synchronized binaries are pairs of stars whose rotation periods are equal to
their orbital period.
Since synchronization appears to happen at rotation periods of 7 days or
shorter \citep{simonian2019}, and most isolated stars have rotation periods
longer than 7 days, the rotation periods of synchronized binaries are likely
to be {\it shorter} than they would be if they were isolated stars.
For this reason, their rotation periods do not reflect their ages and the
gyrochronal age of a synchronized binary is likely to be much younger than the
true age of the system.
Synchronized binaries are therefore a source of contamination for
gyrochronology and should be removed from samples before performing a
gyrochronal age analysis.
Figure \ref{fig:gap} shows that some of the most rapidly rotating stars in the
\mct\ sample have relatively large absolute velocities, indicating that they
are likely synchronized binaries.
% The large velocity dispersions of the most rapidly rotating stars in the \mct\
% sample indicates that at least some fraction of these stars are likely to be
% synchronized binaries.
For this reason, the velocity dispersions of stars with rotation periods
shorter than the lower edge of the rotation period distribution (green
triangles in figure \ref{fig:gap}) are not significantly smaller than the,
presumed older, orange-colored stars.
% This is an indication that many of these are synchronized binaries.
% The rotation periods of stars in synchronized binaries are tidally locked to
% the orbital period of the binary system and since synchronized binaries have
% short orbital periods this can produce old, seemingly isolated stars, with
% short rotation periods.
% Given the strong correlation between binary fraction and primary mass
% \citep[\eg][]{price-whelan2020}, and the fact that M dwarfs take longer to
% transition from the `I` (rapidly rotating) sequence to the `C' (slowly
% rotating) sequence, it is likely that
In general, stars with rotation periods less than $\sim$10 days have an
increased chance of being synchronized binaries.
This result is in agreement with a recent study which found that a large
fraction of photometric binaries were rapid rotators, and the probability of a
star being a synchronized binary system substantially increased below rotation
periods of around 7 days \citep{simonian2019}.
We caution users of rotation period catalogs that rapid rotators with large
absolute velocities should be flagged as potential synchronized binaries
before applying any gyrochronal analysis.

% \subsection{Validating \vb\ dispersion as an age proxy}
% \label{sec:mass-dependent-heating}

% The conclusions drawn in this paper depend on the assumption that velocity
% dispersion in the direction of Galactic latitude (\sigmavb) can be used as an
% age proxy.
% There are two main reasons however, why \vb\ velocity dispersion may {\it not}
% be a good age proxy.
% Firstly, mass-dependent heating may act on the sample, meaning that velocity
% dispersion depends on both age and mass, so cannot be interpreted as a simple
% age proxy.
% Secondly, since stars in the \kepler\ field have a range of Galactic
% latitudes, using \vb\ as a stand-in for \vz\ may not be equally valid for all
% stars, and introduce a velocity bias for high latitude stars (which are more
% likely to be cooler and older).
% In this section we demonstrate that neither of these problems seem to be a
% significant issue for our data.

% In order to establish whether \sigmavb\ can be used as an age proxy, we
% searched for signs of mass-dependent heating within the \kepler\ field.
% Mass-dependent dynamical heating may result from lower-mass stars experiencing
% greater velocity changes when gravitationally perturbed than more massive
% stars.
% It has not been unambiguously observed in the galactic disk because of the
% strong anti-correlation between stellar mass and stellar age.
% Less massive stars do indeed have larger velocity dispersions, however they
% are also older on average.
% This mass-age degeneracy is highly reduced in M dwarfs because their
% main-sequence lifetimes are longer than the age of the Universe, and no
% evidence for mass-dependent heating has previously been found in M dwarfs
% \citep[\eg][]{faherty2009, newton2016}.

% To investigate whether mass-dependent heating could be acting on the \kepler\
% sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
% whose MS lifetimes exceed around 11 Gyr and are therefore representative of
% the initial mass function.
% We could not perform this analysis on the \mct\ sample, because it only
% includes stars with {\it detectable} rotation periods, and since lower-mass
% stars stay active for longer it is likely that it contains a strong mass-age
% correlation.
% We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
% greater than 1.2 (corresponding to an effective temperature $\lesssim$
% 4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
% We also eliminated photometric binaries by removing stars above a 6th order
% polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
% figure \ref{fig:age_gradient}).
% We then applied the quality cuts described above in section
% \ref{sec:the_data}.
% To search for evidence of mass-dependent heating we calculated the (\vb)
% velocity dispersion of stars in effective temperature bins.
% Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
% outliers before calculating the standard deviation of stars in each bin.
% These extreme velocity outliers may be very old late K and M dwarfs, or they
% result from using \vb\ instead of \vz, which introduces additional velocity
% scatter.

% Figure \ref{fig:vb_vs_teff} shows velocity and velocity dispersion as a
% function of effective temperature for the K and M \kepler\ dwarf sample.
% Velocity dispersion very slightly {\it decreases} with decreasing temperature,
% the opposite of the trend expected for mass-dependent heating, however the
% slope is only inconsistent with zero at 1.3 $\sigma$.
% This trend may be due to a selection bias: cooler stars are fainter and
% therefore typically closer, with smaller heights above the galactic plane and
% smaller velocities.
% The essential point however, is that we do not see evidence for mass-dependent
% heating acting on stars in the \kepler\ field, indicating that velocity
% dispersion {\it can} be used as an age proxy (with the caveat that there is
% still a chance, albeit a small one, that the opposing effects of the selection
% function and mass-dependent heating are working to cancel each other out).
% This analysis was performed using \vb\ but we also examined the {\it vertical}
% velocities of the 537 stars in this sample with RV measurements.
% Again, no evidence was found for mass-dependent heating: the slope of the
% velocity dispersion-temperature relation was consistent with zero.
% \begin{figure}
%   \caption{
%       Top: Stellar velocity (\vb) as a function of \teff\ for
%       \kepler\ K and M dwarfs.
% Vertical lines indicate different \teff-groupings used to calculate velocity
%     dispersion.
% Pink stars were not included in velocity dispersion calculations as they were
%     either removed as outliers during a sigma clipping process, or they lie at
%     the sparcely populated, extremely cool end of the temperature range.
%     Velocity dispersion and \teff\ are slightly positively correlated, likely
%     due to a brightness-related selection bias, indicating that mass-dependent
%     heating does not significantly affect low-mass stars in the \kepler\
%     field.
% }
%   \centering
%     \includegraphics[width=1\textwidth]{vb_vs_teff}
% \label{fig:vb_vs_teff}
% \end{figure}

% Having found no strong evidence for mass-dependent heating, we next tested
% the validity of \vb\ as a proxy for \vz\ in more detail.
% At a galactic latitude, $b$, of zero, $v_b=v_z$, however for increasing values
% of $b$, this equivalence becomes an approximation that grows noisier with $b$.
% To test the validity of the \vb$\sim$\vz\ approximation over a range of
% latitudes we downloaded stellar data from the \Gaia\ Universe Model Snapshot
% (GUMS) simulation -- a simulated \Gaia\ catalog \citep{robin2012}.
% We downloaded stars from four pointings in the \kepler\ field with galactic
% latitudes of around 5\degrees, 10\degrees, 15\degrees, and 20\degrees, out to
% a limiting magnitude of 16 dex, and calculated their \vz\ and \vb\ velocities.
% The relationship between \vz\ and \vb\ is close to 1:1, with \vz\ greater than
% \vb\ by around 4.5 kms$^{-1}$ at $b=5$, due to the Sun's own motion in the
% Galaxy.
% We subtracted this offset and examined the residuals of the \vz\ -- \vb\
% relationship to investigate the variance as a function of Galactic latitude
% (shown in figure \ref{fig:vb_vz}).
% We found that \vb\ is drawn from a heavy-tailed distribution, centered on \vz,
% with standard deviation increasing with $b$ (see figure \ref{fig:vb_vz}).
% The standard deviation of \vz-\vb\ was around 3kms$^{-1}$ at $b \sim 5^\circ
% $, 4kms$^{-1}$ at 10$^\circ$, 6kms$^{-1}$ at 15$^\circ$, and 9kms$^{-1}$ at
% 20$^\circ$.
% This demonstrates that using \vb\ instead of \vz\ for stars in the \kepler\
% field will introduce an additional velocity scatter, inflating \sigmavb\
% relative to \sigmavz.
% This additional velocity scatter will be greatest for stars at the highest
% Galactic latitudes.
